Predictions for detecting BHNS and other double compact objects binaries with LISA

### 2021-04-28 paper/tex/sections/methods.tex
more methods comments done

@@

We use a population of binaries recently presented in \citet{Broekgaarden+2021} and Broekgaarden et al. (in prep). This population is synthesised using the rapid population synthesis code \href{https://compas.science}{COMPAS}\footnote{\url{https://compas.science}} \citep[][Team COMPAS: J. Riley et al. (in prep)]{Stevenson+2017, Vigna-Gomez+2018, Stevenson+2019}. COMPAS broadly follows the approach of other codes, such as BSE \citep{Hurley+2002}. Each sampled binary is initialised as a zero-age main sequence (ZAMS) binary and each component star is evolved using fitting formulae from \citet{Hurley+2000} to the single star models of \citet{Pols+1998}.

@@

One million binaries, $N_{\rm binary} \sim 10^6$, were simulated for each of 50 metallicity bins equally spaced in log space with $Z \in [0.0001, 0.022]$, where $Z$ is the mass fraction of heavy elements. These bins span the allowed metallicities range for the original fitting formulae on which COMPAS is based \citep{Hurley+2000}. This is repeated for 14 physics variations (see Section \ref{sec:variation_assumptions}) and so in total 750 million binaries were simulated.

Each binary is sampled from initial distributions for the primary and secondary masses as well as the separation. The primary mass, or the mass of the initially more massive star, is restricted to $m_1 \in [5, 150] \unit{M_{\odot}}$, which spans the range of interest for NS and BH formation in binary systems, and drawn from the \citet{Kroupa+2001} Initial Mass Function (IMF), $p(m_1) \propto m_1^{-2.3}$. The secondary mass is drawn using the mass ratio of the binary, which we assume to be uniform on $[0, 1]$, therefore $p(q) = 1$ \citep[consistent e.g. with][]{Sana+2012}. We additionally restrict $m_2 \ge 0.1 \unit{M_{\odot}}$, since this is approximately the minimal mass for a main sequence star. We assume that the initial separation follows a flat in the log distribution with $p(a_i) \propto 1 / a_i$ and $a_i \in [0.01, 1000] \unit{AU}$ \citep{Opik+1924, Abt+1983}. We assume that all binaries are circular at birth to reduce the dimensions of initial parameters. Since we focus on post-interaction binaries which will have circularised during mass transfer this is a reasonable assumption and is likely not critical for predicting detection rates \citep{Hurley+2002, deMink+2015}. In addition, we apply the adaptive importance sampling algorithm STROOPWAFEL \citep{Broekgaarden+2019} to improve the yield of our sample. This algorithm increases the prevalence of target DCOs (BHBHs, BHNSs and NSNSs in this case) in the sample and assigns each a weight, $w$, which represents the probability of drawing it without STROOPWAFEL in effect.

For each metallicity $Z$, we have a sample of binaries, each with a set of parameters
\begin{equation}
    \mathbf{b}_{{Z, i}} = \{m_1, m_2, a_{\rm DCO}, e_{\rm DCO}, t_{\rm evolve}, t_{\rm inspiral}, w\},
\end{equation}
for $i = 1, 2, \dots, N_{\rm binary}$, where $m_1$ and $m_2$ are the primary and secondary masses, $a_{\rm DCO}$ and $e_{\rm DCO}$ are the semi-major axis and eccentricity at the moment of double compact object (DCO) formation, $t_{\rm evolve}$ is the time between the binary's zero-age main sequence and DCO formation, $t_{\rm inspiral}$ is the time between DCO formation (that is immediately after the second supernova in the system) and gravitational wave merger, $w$ is the adaptive importance sampling weight assigned by STROOPWAFEL. We sample from these sets of parameters when creating synthetic galaxies.

\subsubsection{Physical assumptions in fiducial model}\label{sec:fiducial_physics}
In this section we briefly summarise the main physical assumptions in our fiducial model. For more details see Section 2.1.2 and Table 1 of \citet{Broekgaarden+2021}.

@@

\subsubsection{Model variations} \label{sec:variation_assumptions}
In addition to our fiducial model for the formation of DCOs, we explore 14 other models in which we change various aspects of the mass transfer, common envelope and supernova physics assumptions in order to assess their effect on the overall double compact object detection rates and distributions. Each of the models varies a single physics assumption (fiducial assumptions are outlined in Section~\ref{sec:fiducial_physics}) and these are outlined in Table~\ref{tab:physics_variations}.

Our fiducial model is labelled model \modFid{}. Models \modRangeMT{} focus on changes to the mass transfer physics. We explore the effect of fixing the mass transfer efficiency $\beta$ to a constant value, rather than allowing it to vary based on the maximum accretion rate, in models \modBetaLow{}, \modBetaMed{}, \modBetaHigh{}, in which we set the value of $\beta$ to $0.25$, $0.5$ and $0.75$ respectively. In model \modCaseBB{} we investigate the consequence of assuming that case BB mass transfer onto a neutron star or black hole is always stable rather than always unstable.

@@

Finally, models \modRangeSN{} make changes related to our assumptions about supernova physics. Model \modRapid{} uses the alternate \textit{rapid} remnant mass prescription from \citet{Fryer+2012} instead of the \textit{delayed} prescription. We change the maximum neutron star mass in models \modNSLow{} and \modNSHigh{} to $2$ and $3 \unit{M_{\odot}}$ respectively to account for the range of predicted maximum neutron star masses. Model \modNoPISN{} removes the implementation of pair-instability and pulsational pair-instability supernovae. In models \modSigLow{} and \modSigLower{} we decrease the root-mean-square velocity dispersion for core-collapse supernovae to explore the effect of lower kicks. Finally, model \modNoBH{} removes the natal kick for all black holes.

@@

\subsection{Galaxy synthesis}

@@

\subsection{Gravitational wave detection}
We use the Python package \href{https://legwork.readthedocs.io/en/latest/}{LEGWORK} to evolve binaries and calculate their LISA detectability. For a full derivation of the equations given below please see the LEGWORK release paper (Wagg et al. in prep) or \href{https://legwork.readthedocs.io/en/latest/notebooks/Derivations.html}{documentation}.

@@

\subsubsection{Detection rate calculation}
For each physics variation model, we compute the total number of merging DCOs in the Milky Way today based on the COMPAS simulation. This takes into account the true mass and IMF of the Milky Way since our simulations each have a slightly smaller galaxy size and only sample from massive stars rather than the full IMF. For a more in-depth discussion of this process see Appendix~\ref{app:rate_normalisation}.

We determine the fraction of binaries that are detectable in each Milky Way instance by summing the adaptive importance sampling weights of the binaries that have an SNR greater than 7 and dividing by the total weights in the simulation. We multiply this fraction by the total number in the Milky Way to find a detection rate.
\begin{equation}
    N_{\rm detect} = \frac{\sum_{i = 0}^{N_{\rm detect}} w_i}{\sum_{i = 0}^{N_{\rm DCO}} w_i} \cdot N_{\rm MW}
\end{equation}
We calculate this detection rate by Monte Carlo sampling 2500 Milky Way instances (each containing 100,000 DCOs) for each DCO type and every physics variation in order to obtain values for the uncertainty on the expected detection rate.

>>>>

To produce predictions for the DCOs that are detectable with LISA, we synthesise a population of DCOs using the population synthesis methods described in Section~\ref{sec:COMPAS_explained}. In order to obtain value for the uncertainty on the expected detection rates, we place this sample of DCOS in many different Monte Carlo sampled instances of the Milky Way, the model for which is described in Section~\ref{sec:galaxy_synthesis}. We evolve the orbit of each DCO in a Milky Way instance up to the LISA mission and calculate the detection rate for that instance using the methods presented in Section~\ref{sec:gw_detection}.

@@

We use a population of binaries recently presented in \citet{Broekgaarden+2021} and Broekgaarden et al. (in prep). This population is synthesised using the rapid population synthesis code \href{https://compas.science}{COMPAS}\footnote{\url{https://compas.science}} \citep{Stevenson+2017, Vigna-Gomez+2018, Stevenson+2019}. COMPAS follows the approach of the pioneering population synthesis code BSE \citep{Hurley+2000,Hurley+2002} and uses fitting formula and rapid algorithms to efficiently predict the final fate of millions of binary systems. The code is open source and documented in the papers listed above and the on-line documentation. An extensive method paper is forthcoming (Team COMPAS: J. Riley et al. (in prep)) We summarize the main assumptions and settings relevant for this work in the following sections.

@@

One million binaries are simulated for 50 metallicity bins equally spaced in log space between $Z \in [0.0001, 0.022]$, where $Z$ is the mass fraction of heavy elements. These bins span the allowed metallicities range for the original fitting formulae on which COMPAS is based \citep{Hurley+2000}. This is repeated for 14 physics variations (see Section \ref{sec:variation_assumptions}) and so in total 750 million binaries were simulated.

Each binary is sampled from initial distributions for the primary and secondary masses as well as the separation. The primary mass, that is the mass of the initially more massive star, is restricted to $m_1 \in [5, 150] \unit{M_{\odot}}$, which spans the range of interest for NS and BH formation in binary systems, and drawn from the \citet{Kroupa+2001} Initial Mass Function (IMF), $p(m_1) \propto m_1^{-2.3}$. The secondary mass is drawn using the mass ratio of the binary, which we assume to be uniform on $[0, 1]$, therefore $p(q) = 1$ \citep[consistent e.g. with][]{Sana+2012}. We additionally restrict $m_2 \ge 0.1 \unit{M_{\odot}}$, since this is approximately the minimal mass for a main sequence star. We assume that the initial separation follows a flat in the log distribution with $p(a_i) \propto 1 / a_i$ and $a_i \in [0.01, 1000] \unit{AU}$ \citep{Opik+1924, Abt+1983}. We assume that all binaries are circular at birth to reduce the dimensions of initial parameters. Since we focus on post-interaction binaries which will have circularised during mass transfer this is a reasonable assumption and is likely not critical for predicting detection rates \citep{Hurley+2002, deMink+2015}. In addition, we apply the adaptive importance sampling algorithm STROOPWAFEL \citep{Broekgaarden+2019} to improve the yield of our sample. This algorithm increases the prevalence of target DCOs (BHBHs, BHNSs and NSNSs in this case) in the sample and assigns each a weight, $w$, which represents the probability of drawing it without STROOPWAFEL in effect.

For each metallicity $Z$, we thus have a sample of binaries, each with a set of parameters
\begin{equation}
    \mathbf{b}_{{Z, i}} = \{m_1, m_2, a_{\rm DCO}, e_{\rm DCO}, t_{\rm evolve}, t_{\rm inspiral}, w\},
\end{equation}
for $i = 1, 2, \dots, N_{\rm binary}$, where $m_1$ and $m_2$ are the primary and secondary masses, $a_{\rm DCO}$ and $e_{\rm DCO}$ are the semi-major axis and eccentricity at the moment of double compact object (DCO) formation, $t_{\rm evolve}$ is the time between the binary's zero-age main sequence and DCO formation, $t_{\rm inspiral}$ is the time between DCO formation (that is immediately after the second supernova in the system) and gravitational wave merger, $w$ is the adaptive importance sampling weight assigned by STROOPWAFEL. We sample from these sets of parameters when creating synthetic galaxies.

\subsubsection{Physical assumptions in fiducial model}\label{sec:fiducial_physics}
In this section we briefly summarise the main physical assumptions in our fiducial model. For more details see \citet{Broekgaarden+2021}.

\textit{Stellar Evolution:} To follow the evolution of massive stars, COMPAS relies on fitting formula by \citet{Hurley+2000} to detailed single star models by \citet{Pols+1998}. COMPAS implements stellar wind mass loss using the prescriptions from \citet{Belczynski+2010b} and models the evolution of stars that lose or gain mass closely following the algorithms originally described in \citet{Tout+1996} and \citet{Hurley+2002}. For more information about the implementation of the single star evolution in COMPAS see the corresponding section in the upcoming methods paper (Team COMPAS: J. Riley et al. (in prep)).

@@

\subsubsection{Model variations} \label{sec:variation_assumptions}
In addition to our fiducial model for the formation of DCOs, we explore 14 other models in which we change various aspects of the mass transfer, common envelope and supernova physics assumptions in order to assess the effect of their uncertainties on the overall double compact object detection rates and distributions. Each of the models varies a single physics assumption (fiducial assumptions are outlined in Section~\ref{sec:fiducial_physics}) and these are outlined in Table~\ref{tab:physics_variations}.

Our fiducial model is labelled model \modFid{}. Models \modRangeMT{} focus on changes to the mass transfer physics assumptions. We explore the effect of fixing the mass transfer efficiency $\beta$ to a constant value, rather than allowing it to vary based on the maximum accretion rate. In models \modBetaLow{}, \modBetaMed{}, \modBetaHigh{}, in which we set the value of $\beta$ to $0.25$, $0.5$ and $0.75$ respectively. In model \modCaseBB{} we investigate the consequence of assuming that case BB mass transfer onto a neutron star or black hole is always stable rather than always unstable.

@@

Finally, in models \modRangeSN{} we consider changes related to our assumptions about supernova physics. Model \modRapid{} uses the alternate \textit{rapid} remnant mass prescription from \citet{Fryer+2012} instead of the \textit{delayed} prescription. We change the maximum neutron star mass in models \modNSLow{} and \modNSHigh{} to $2$ and $3 \unit{M_{\odot}}$ respectively to account for the range of predicted maximum neutron star masses. Model \modNoPISN{} removes the implementation of pair-instability and pulsational pair-instability supernovae. In models \modSigLow{} and \modSigLower{} we decrease the root-mean-square velocity dispersion for core-collapse supernovae to explore the effect of lower kicks. Finally, model \modNoBH{} removes the natal kick for all black holes.

@@

\subsection{Galaxy synthesis}\label{sec:galaxy_synthesis}

@@

\subsection{Gravitational wave detection}\label{sec:gw_detection}
We use the Python package \href{https://legwork.readthedocs.io/en/latest/}{LEGWORK} to evolve binaries and calculate their LISA detectability. For a full derivation of the equations given below please see the LEGWORK release paper (Wagg et al. in prep) or \href{https://legwork.readthedocs.io/en/latest/notebooks/Derivations.html}{documentation}.

@@

\subsubsection{Detection rate calculation}
For any one instance of the Milky Way, we calculate the detection rate

For each physics variation model, we compute the total number of merging DCOs in the Milky Way today, $N_{\rm MW}$, based on the COMPAS simulation. This takes into account the true mass and IMF of the Milky Way since our simulations each have a slightly smaller galaxy size and only sample from massive stars rather than the full IMF. For a more in-depth discussion of this process see Appendix~\ref{app:rate_normalisation}.

We determine the fraction of binaries that are detectable in each Milky Way instance by summing the adaptive importance sampling weights of the binaries that have an SNR greater than 7 and dividing by the total weights in the simulation. We multiply this fraction by the total number in the Milky Way to find a detection rate.
\begin{equation}
    N_{\rm detect} = \frac{\sum_{i = 0}^{N_{\rm detect}} w_i}{\sum_{i = 0}^{N_{\rm DCO}} w_i} \cdot N_{\rm MW}
\end{equation}
We calculate the detection rate by Monte Carlo sampling 2500 Milky Way instances (each containing 100,000 DCOs) for each DCO type and every physics variation in order to obtain values for the uncertainty on the expected detection rate.
